Statically Typed Web Programming using Scala

### 2012-12-14 papers/ecoop2013/repairing.tex
Weekend

\author{Julien \noun{Richard-Foy}\textsuperscript{\textasteriskcentered}\textsuperscript{\textdaggerdbl} \and Olivier
\noun{Barais}\textsuperscript{\textasteriskcentered} \and Tiark \noun{Rompf}\textsuperscript{\textdagger} \and
Jean-Marc \noun{Jézéquel}\textsuperscript{\textasteriskcentered}}
\institute{\textsuperscript{\textasteriskcentered}IRISA, Université de Rennes 1, France:
\texttt{\{first\}.\{last\}@irisa.fr}\\
\textsuperscript{\textdaggerdbl}Zenexity, Paris, France: \texttt{jrf@zenexity.com}\\
\textsuperscript{\textdagger}EPFL, Lausanne, Switzerland: \texttt{\{first\}.\{last\}@epfl.ch}}

@@

The Web is an appealing platform on which to write applications (\aka{} Rich Internet Applications, RIA) because it
makes them easy to deploy on clients and allows large scale innovative collaborative
experiences~\cite{Farrell07_RIA,Mikkonen08_SpaghettiJs}. RIAs are characterized by partial updates of the user
interface (as opposed to refreshing the whole page with the classic hyperlink navigation), so a large part of the
client-side code listens to user interface events (such as mouse clicks), triggers the appropriate action
asynchronously on the server using AJAX~\cite{Garrett05_Ajax} and updates the application’s state and the DOM content
with a new DOM fragment computed from the response data~\cite{Farrell07_RIA,Busch09_StateOfArtRIA}.

Writing large Web applications is known to be
challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous
nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
writing distributed code requires its different parts to be consistent and leads to code duplication unless one uses
the same language in both sides, which, in the case of the Web, means using JavaScript to write the whole
application. However this language has several drawbacks making it hardly suitable for large code base (\eg{} no
static typing, no module system, verbose syntax, \etc). Some other in-browser execution environments give the
opportunity to write the client-side code in another language that JavaScript, \eg{} Java
applets~\cite{Anuff96_JavaApplets}, Adobe Flash~\cite{Curtis00_Flash} and Microsoft
Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However
these technologies also have several drawbacks: they require an additional browser plugin to be installed (which may
not be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
Apple smart phone), the page content can't naturally be referenced by search engines, and the content is not
structured in URLs that users can bookmark or share.

@@

This paper continues their work and presents safe and expressive DSL constructs for RIAs development. We increase the
expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed APIs. We
make asynchronous programming simpler by raising asynchronous computations to first-class citizen. We leverage the
Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and type coercion to
the client-side.

@@

The next section introduces a simple application example showing challenges of RIA development, section
\ref{solution} presents our DSL tackling these challenges, section \ref{validation} validates our solution, section
\ref{related} presents related works and compares it to related work and section \ref{discussion} concludes.

@@

We claim this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to
mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to
the listened event but developers have no way to check that the fields they’re using are indeed defined on the event
they’re listening to. For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the
\code{mousewheel} event.

@@

Another issue with callback-driven programming arises when several dependent asynchronous computations are executed
sequentially. Consider for example the following listing performing three consecutive Ajax requests:

\begin{lstlisting}[language=JavaScript,label=async-callback-hell,caption=Sequential asynchronous calls]
Ajax.get(fooUrl(), function (foo) {
  Ajax.get(barUrl(foo), function (bar) {
    Ajax.get(bazUrl(bar), function (baz) {
      console.log(foo + bar + baz);
    });
    console.log("bar = " + bar)
  });
});
\end{lstlisting}

Notice that the code is getting deeper toward the right, this problem is often referred to as the
“callback-hell”~\cite{McKenna_Roy}. This code is also hard to reason about: when will the second \code{console.log}
statement be executed? Before or after \code{baz} has been fetched?

@@

\begin{lstlisting}[language=JavaScript,label=dom-api,caption=DOM fragment creation using the native API]
var vertexDom = function (v) {
  var root = document.createElement("g");
  root.setAttribute("class", "vertex");
  root.setAttribute("transform",
    "translate(" + v.x + "," + v.y + ")");
  var rect = document.createElement("rect");
  rect.setAttribute("width", v.width);
  rect.setAttribute("height", v.height);
  var text = document.createElement("text");
  text.setAttribute("width", v.width);
  text.setAttribute("height", v.height);
  text.appendChild(document.createTextNode(v.content));
  root.appendChild(rect);
  root.appendChild(text);
  return root
};
\end{lstlisting}

@@

The above code may be more readable, however this implementation is wrong: if we call it with a content containing
brackets (\eg{} \code{"<foo>"}) they won’t be escaped and will produce a \code{foo} tag nested in the \code{text} tag,
which is not the intended behavior. So, although this way reads slightly better it’s not less error prone.

@@

HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or
Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}}
aim to make it simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML
structure and allowing the insertion of dynamic expressions in a safe way. Some template engines can be used on both
client-side and server-side, however none has a practical expression language (dynamic content can only be provided
as a map of key-value pairs and no operation can be done on values) or is statically typed.

@@

In the previous section we showed the three main tasks performed by the JavaScript code in RIAs: events handling,
asynchronous programming and DOM manipulation. We noticed the following issues:

\subsubsection{Lack of static hecks}

@@

The lack of static checks is even harder to tackle in the case of code shared between client and server sides. If we
write the expression \code{xs.size == 0} in a template, how should it be translated to the client-side environment
(JavaScript) and to the server-side environment (\eg{} the Java Virtual Machine)? Several solutions are valid,
depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or
\code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a
collection is named \code{length}). We need to define a proper expression language and to define how it should be
translated for the client-side environment and the server-side environment, but such a task requires a high effort
and is hard to achieve in an extensible way (so users can still integrate their own data types in expressions instead
of being restricted to a set of supported types).

@@

JavaScript libraries could address some of the challenges presented above, but addressing type safety and
asynchronous programming issues require language modifications (\eg{} to bring a type system and a sequencing
notation). Defining a language requires a high effort, especially to build development tools for the language so we
chose to define our language as a compiled embedded DSL in Scala\footnote{The code is available at
\href{http://github.com/js-scala}{http://github.com/js-scala}}.

@@

We use the Lightweight Modular Staging framework (LMS) to define our compiled embedded DSL. The main idea is that a
program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate
representation of the program and the second stage transforms this intermediate representation into executable code
(figure \ref{lms-diagram}). The embedding lets us reuse the host language infrastructure (compiler, development
tools) and type system, and staging gives the opportunity to generate efficient code from expressive code by applying
domain specific optimizations.

@@

We want to bring safety to the event handling API so that developers cannot mispell an event name and cannot attempt
to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
type of the event passed to the handler varies with the event name. For instance in the listing \ref{event-js}, the
handler processses \code{MouseWheelEvent} values because it listens to \code{mousewheel} events. Values of type
\code{MouseWheelEvent} have a property \code{wheelDeltaY} but that’s not the case of an event data of type
\code{KeyboardEvent}, for example.

@@

\begin{lstlisting}
class EventDef {
  type Type
}

def on(ev: EventDef)(handler: Rep[ev.Type] => Rep[Unit]): Rep[Unit]
\end{lstlisting}

(We renamed \code{addEventListener} to \code{on}, for the sake of brevity). An \code{EventDef} value represents an
event that carries its corresponding event data type in its \code{Type} type member. The \code{on} method takes a
parameter \code{ev} of type \code{EventDef} and a \code{handler} parameter whose type refers to the \code{Type}
member of the \code{ev} parameter, so the type of the handler depends on the \code{ev} value. We can then define the
\code{keyup} and \code{mousewheel} events as follows:

\begin{lstlisting}
object KeyUp extends EventDef { type Type = KeyboardEvent }
object MouseWheel extends EventDef { type Type = MouseWheelEvent }
\end{lstlisting}

@@

We propose a DSL that explicitly reflects the asynchronous nature of computations in their types and provides methods
turning them into first-class citizens. The DSL is monadic so we can solve the callback-hell problem thanks to the
Scala \code{for} notation.

For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be
re-written with our DSL. The \code{createVertex} now returns an asynchronous value instead of taking a callback as
parameter. Then, the \code{for} expression allows us to use the vertex, when available, and to insert it on the user
interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
a parameter, the code is easier to modularize into loosely coupled parts.

@@

The listing \ref{async-no-callback} translates the “callback-hell” example (listing \ref{async-callback-hell}) using
our DSL. The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the
first Ajax request is available, the next statement will be executed, and so on. There is no nested callbacks and the
order of execution is directly reflected by the order of statements.

@@

In this section we show how we can define a template engine as an embedded DSL, with minimal effort. This template
engine is statically typed and able to insert dynamic content in a safe way. It provides a powerful expression
language, requires no extra compilation step and can be used on both client-side and server-side.

@@

So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
function to define a tag and a \code{text} function to define a text node. The following listing uses our DSL and
generates a code equivalent to listing \ref{dom-api}:

\begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
def vertexDom(v: Rep[Vertex]) =
    tag("g", "class"->"vertex",
             "transform"->("translate("+v.x+","+v.y+")"))(
        tag("rect", "width"->v.width, "height"->v.height)(),
        tag("text", "width"->v.width, "height"->v.height)(
            text(v.content)
        )
    )
\end{lstlisting}

The readability has been highly improved: nesting tags is just like nesting code blocks, HTML entities are
automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.

@@

Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
client sides is surprisingly as simple as defining another code generator for the DSL, producing Scala code.

@@

Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to call a
specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
\emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
the objects. However, modifying existing object prototypes is considered bad
practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
registering supported data types at the beginning of the program execution, as described in section 2.4.3
of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.

@@

In summary, the compiled embedded approach may require more work to support custom data types but gives control on
the way this data type will be encoded for each target platform (\eg{} the JavaScript code generator uses native
arrays to encode the lists of the list manipulation DSL).

@@

Furthermore, a code generator can be tweaked by defining a sub-trait specializing some of its methods. This is useful
to handle cross-browser incompatibilities. For instance we have written code generators specializing the output for
some statement nodes in order to handle Internet Explorer incompatibilities.

@@

This section presents relevant implementation details of the js-scala DSL.

@@

The \emph{ad hoc} polymorphism feature of our DSL is directly borrowed from Scala’s typeclasses but the type coercion
feature required to define a type-level DSL to express constraints. Numeric operations take an implicit value of
type \code{(A \~ B) ~> C}. This type is actually a syntactic sugar for the type \code{\~>[\~[A, B], C]} (Scala
supports an infix notation for type constructors of arity two). The \code{\~\>} type also takes two value parameters,
saying how to obtain a value of the target type for each operand. Here is the definition of the \code{\~\>} type:

@@

  def option_isEmpty[A](o: Rep[Option[A]]): Rep[Boolean]
  def option_foreach[A](o: Rep[Option[A]], f: Rep[A] => Rep[Unit]): Rep[Unit]
}
// Intermediate representation
trait OptionOpsExp extends OptionOps { this: EffectExp =>
  def option_isEmpty[A](o: Exp[Option[A]]) = OptionIsEmpty(o)
  def option_foreach[A](o: Exp[Option[A]], f: Exp[A] => Exp[Unit]) = {
    val a = fresh[A]
    val block = reifyEffects(f(a))
    reflectEffect(OptionForeach(o, a, block), summarizeEffects(block).star)
  }

@@

We limited the DSL to two operations for the sake of brevity. The \code{OptionOps} trait defines the user facing DSL,
it enriches \code{Rep[Option[_]]} values by providing an implicit conversion to \code{OptionOpsCls} that defines the
\code{isEmpty} and \code{foreach} methods that respectively tests the emptiness of a value and derefences a value if
not empty. These methods provide a convenient syntax to use \code{Rep[Option[_]]} values and delegate to the
\code{option_isEmpty} and \code{option_foreach} operations, respectively, that are left abstract. The
\code{OptionOpsExp} trait implements the \code{option_isEmpty} and \code{option_foreach} methods.
The \code{option_isEmpty} implementation simply returns a statement node of type \code{ForeachIsEmpty}, but the
\code{option_foreach} implementation is more subtle because it has to track the dependencies of its function
parameter \code{f} (for instance, side-effects performed within the \code{f} function must be reflected by the
returned \code{OptionForeach} node).

The principle is the same for all DSLs: first, provide a trait with the DSL vocabulary, then implement each DSL
construct to return a reified intermediate representation of itself. However, sometimes you have the opportunity to
perform some optimizations at the evaluation level such as constants folding. For instance, the DOM definition DSL
builds trees consisting of a root node and eventually several children. If these children are known at evaluation
time we can build a node having a \code{List[Rep[Node]]} value as children instead of a \code{Rep[List[Node]]} so the
generated code will be able to inline the loop adding the children to the parent node. Listing \ref{forest-impl}
shows the relevant part of code checking if the \code{children} parameter of the \code{forest_tag} method is a
constant list (the \code{ListNew} statement node represents constant lists) and building different intermediate
representations according to this test.

\begin{figure}
\begin{lstlisting}[caption=Optimized intermediate representation for the DOM definition DSL,label=forest-impl]
trait ForestExp extends Forest with EffectExp { this: ListOpsExp =>
  def forest_tag(name: String, attrs: Map[String, Exp[String]], children: Exp[List[Node]]) = {
    reflectEffect {
      children match {
        case Def(ListNew(children)) =>
          Tag(name, Left(children.toList), attrs)
        case _ =>
          Tag(name, Right(children), attrs)
      }
    }
  }
  case class Tag(name: String, children: Either[List[Exp[Node]], Exp[List[Node]]], attrs: Map[String, Exp[String]]) extends Def[Node]
}
\end{lstlisting}
\end{figure}

@@

Code generation consists in traversing the statement nodes procudes by the program evaluation according to their
dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
authors just need to say how to emit code for each statement node of their DSL. Listing \ref{option-codegen} shows
such a code generator for the \code{null} reference handling DSL. The \code{emitNode} method handles
\code{OptionIsEmpty} and \code{OptionForeach} nodes. In the case of the \code{OptionIsEmpty} node, it simply
generates an expression testing if the value is \code{null}, in the case of the \code{OptionForeach} node, it wraps
the code block using the value within a \code{if} checking that the value is not \code{null}.

>>>>

\author{Julien \noun{Richard-Foy}\textsuperscript{\textasteriskcentered}\textsuperscript{\textdaggerdbl} \and Olivier
\noun{Barais}\textsuperscript{\textasteriskcentered} \and Tiark \noun{Rompf}\textsuperscript{\textdagger} \and
Jean-Marc \noun{Jézéquel}\textsuperscript{\textasteriskcentered}}
\institute{\textsuperscript{\textasteriskcentered}IRISA, Université de Rennes 1, France:
\texttt{\{first\}.\{last\}@irisa.fr}\\
\textsuperscript{\textdaggerdbl}Zenexity, Paris, France: \texttt{jrf@zenexity.com}\\
\textsuperscript{\textdagger}EPFL, Lausanne, Switzerland: \texttt{tiark.rompf@epfl.ch}}

@@

The Web is an appealing platform on which to write applications (\aka{} Rich Internet Applications, RIA) because it
makes them easy to deploy on clients and allows large scale innovative collaborative
experiences~\cite{Farrell07_RIA,Mikkonen08_SpaghettiJs}. RIAs are characterized by partial updates of the user
interface (as opposed to refreshing the whole page with the classic hyperlink navigation), so a large part of the
client-side code listens to user interface events (such as mouse clicks), triggers the appropriate action
asynchronously on the server using AJAX~\cite{Garrett05_Ajax} and updates the application’s state and the page content
with a new DOM fragment computed from the response data~\cite{Farrell07_RIA,Busch09_StateOfArtRIA}.

Writing large Web applications is known to be
challenging~\cite{Mikkonen08_SpaghettiJs,Preciado05_RIAMethodologyNecessity}, mainly because of the heterogeneous
nature of the client-side and server-side environments~\cite{Echeverria09_RIA,Kuuskeri09_PartitioningClientServer}:
writing distributed code requires its different parts to be consistent and leads to code duplication unless one uses
the same language in both sides, which, in the case of the Web, means using JavaScript to write the whole
application. However this language has several drawbacks making it hardly suitable for large code bases (\eg{} no
static typing, no module system, verbose syntax, \etc). Some other in-browser execution environments give the
opportunity to write the client-side code in another language that JavaScript, \eg{} Java
applets~\cite{Anuff96_JavaApplets}, Adobe Flash~\cite{Curtis00_Flash} and Microsoft
Silverlight\footnote{\href{http://www.microsoft.com/silverlight/}{http://www.microsoft.com/silverlight/}}. However
these technologies also have several drawbacks: they require an additional browser plugin to be installed (which may
not be available on all devices having a Web browser: for instance there is no way to execute Flash objects within an
Apple smart phone), the page content cannot naturally be referenced by search engines, and the content is not
structured in URLs that users can bookmark or share.

@@

This paper continues their work and presents several safe and expressive DSL constructs for RIAs development. We
increase the expressivity of JavaScript native APIs with no runtime overhead and expose them through statically typed
APIs. We make asynchronous programming simpler by raising asynchronous computations to first-class citizens. We
leverage the Scala type system to bring type-directed language features such as \emph{ad hoc} polymorphism and type
coercion to the client-side.

@@

The next section introduces a simple application example showing the main challenges of RIA development, section
\ref{solution} presents our DSL tackling these challenges, section \ref{validation} evaluates our solution, section
\ref{related} compares our solution to related works and section \ref{discussion} concludes.

@@

We claim this code is fragile for two reasons. (1) The name of the event is passed as a string, so it is easy to
mispell it. (2) The callback passed as a second parameter takes a parameter \code{e} whose fields vary according to
the listened event, but developers have no way to check that the fields they’re using are indeed defined on the event
they’re listening to. For instance, in our case we use the \code{wheelDeltaY} event field that is defined only on the
\code{mousewheel} event.

@@

Another issue with callback-driven programming arises when several dependent asynchronous computations are executed
sequentially. Consider for example the code of listing \ref{async-callback-hell} performing three consecutive Ajax
requests. Notice that the code is getting deeper toward the right. This problem is often referred to as the
“callback hell”~\cite{McKenna_Roy}. This code is also hard to reason about: when will the second \code{console.log}
statement be executed? Before or after \code{baz} has been fetched?

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=async-callback-hell,caption=Sequential asynchronous calls]
Ajax.get(fooUrl(), function (foo) {
  Ajax.get(barUrl(foo), function (bar) {
    Ajax.get(bazUrl(bar), function (baz) {
      console.log(foo + bar + baz);
    });
    console.log("bar = " + bar)
  });
});
\end{lstlisting}
\end{figure}

@@

\begin{lstlisting}[language=JavaScript,label=dom-api,caption=DOM fragment creation using the native API]
var vertexDom = function (v) {
  var root = document.createElement("g");
  root.setAttribute("class", "vertex");
  root.setAttribute("transform",
    "translate(" + v.x + "," + v.y + ")");
  var rect = document.createElement("rect");
  rect.setAttribute("width", v.width);
  rect.setAttribute("height", v.height);
  var text = document.createElement("text");
  text.setAttribute("width", v.width);
  text.setAttribute("height", v.height);
  text.appendChild(document.createTextNode(v.content));
  root.appendChild(rect);
  root.appendChild(text);
  return root
};
\end{lstlisting}
% 
% \begin{lstlisting}
% var showMap = function (map) {
% 
% var vertices = function (vs) {
%   return vs.map(function (v) {
%     var g = document.createElementNS('g', 'svgns');
%     g.setAttribute('class', 'vertex');
%     g.setAttribute('transform', "translate(" + v.x + "," + v.y + ")");
%     var rect = document.createElementNS('rect', 'svgns');
%     rect.setAttribute('width', v.width);
%     rect.setAttribute('height', v.height);
%     var f = document.createElement('foreignObject');
%     f.setAttribute('x', 0);
%     f.setAttribute('y', 0);
%     f.setAttribute('width', v.width);
%     f.setAttribute('height', v.height);
%     var p = document.createElement('p');
%     p.setAttribute('style', 'height: '+v.height+'px;');
%     p.appendChild(document.createTextNode(v.content));
%     f.appendChild(p)
%     g.appendChild(rect);
%     g.appendChild(f);
%     return g
%   });
% };
% 
% var edges = function (es) {
%   return es.map(function (e) {
%     var x1 = e.orig.x + e.orig.width / 2;
%     var y1 = e.orig.y + e.orig.height / 2;
%     var x2 = e.end.x + e.end.with / 2;
%     var y2 = e.end.y + e.end.height / 2;
%     var g = document.createElementNS('g', 'svgns');
%     g.setAttribute('class', 'edge');
%     var line = document.createElement('line', 'svg');
%     line.setAttribute('x1', x1);
%     line.setAttribute('x2', x2);
%     line.setAttribute('x3', x3);
%     line.setAttribute('x4', x4);
%     return g;
%   });
% };
% 
% return vertices(map.vertices).concat(edges(map.edges));
% 
% };
% \end{lstlisting}

@@

The above code may be more readable, however this implementation is wrong: if we call it with a content containing
brackets (\eg{} \code{"<foo>"}), they won’t be escaped and will produce a \code{foo} tag nested in the \code{text}
tag, which is not the intended behavior. So, although this way reads slightly better it’s not less error prone.

@@

HTML template engines like Mustache\footnote{\href{http://mustache.github.com/}{http://mustache.github.com/}} or
Closure Templates\footnote{\href{https://developers.google.com/closure/templates/}{https://developers.google.com/closure/templates/}}
aim to make it simpler the definition of DOM fragments by providing a convenient syntax to describe the HTML
structure and allowing the insertion of dynamic expressions in a safe way. Some template engines can be used on both
client-side and server-side, however none is statically typed or has a practical expression language: dynamic content
can only be provided as a map of key-value pairs (rather than typed values) and cannot be processed for
presentational purposes from within the template.

@@

In the previous section we presented the three main tasks performed by the JavaScript code in RIAs: event handling,
asynchronous programming and DOM manipulation. We noticed the following issues:

\subsubsection{Lack of static checks}

@@

The lack of static checks is even harder to tackle in the case of code shared between client and server sides. If we
write the expression \code{xs.size == 0}, how should it be translated to the client-side environment
(JavaScript) and to the server-side environment (\eg{} the Java Virtual Machine)? Several solutions are valid,
depending on the meaning of the expression: in JavaScript we could translate it as \code{xs.size === 0} or
\code{xs.length === 0} if \code{xs} refers to a collection (in JavaScript the property to get the size of a
collection is named \code{length}). So we need to define a proper expression language and to define how it should be
translated for the client-side environment and the server-side environment, but such a task requires a high effort
and is hard to achieve in an extensible way (so users can still integrate their own data types in expressions instead
of being restricted to a set of supported types).

@@

JavaScript libraries could address some of the challenges presented above, but addressing type safety and
asynchronous programming issues require language modifications (\eg{} to bring a type system and a sequencing
notation). Defining a language requires a high effort, especially to build development tools for the language so we
chose to define our language as a compiled embedded DSL in Scala.

@@

We use the Lightweight Modular Staging framework (LMS) to define our compiled embedded DSL. The main idea is that a
program written using a DSL is evaluated in two stages (or steps): the first stage builds an intermediate
representation of the program and the second stage transforms this intermediate representation into executable code
(figure \ref{lms-diagram}).

@@

We want to bring safety to the event handling API so that developers cannot mispell an event name and cannot attempt
to read a property that is not defined on the type of the handled event. The difficulty comes from the fact that the
type of the event passed to the handler varies with the event name. For instance, in listing \ref{event-js} the
handler processes \code{MouseWheelEvent} values because it listens to \code{mousewheel} events. Values of type
\code{MouseWheelEvent} have a property \code{wheelDeltaY} but that’s not the case of an event data of type
\code{KeyboardEvent}, for example.

@@

\begin{lstlisting}
class EventDef {
  type Data
}

def on(ev: EventDef)(handler: Rep[ev.Data] => Rep[Unit]): Rep[Unit]
\end{lstlisting}

(We renamed \code{addEventListener} to \code{on}, for the sake of brevity). An \code{EventDef} value represents an
event that carries its corresponding event data type in its \code{Data} type member. The \code{on} method takes a
parameter \code{ev} of type \code{EventDef} and a \code{handler} parameter whose type refers to the \code{Data}
member of the \code{ev} parameter, so the type of the handler depends on the \code{ev} value. We can then define the
\code{keyup} and \code{mousewheel} events as follows:

\begin{lstlisting}
object KeyUp extends EventDef { type Data = KeyboardEvent }
object MouseWheel extends EventDef { type Data = MouseWheelEvent }
\end{lstlisting}

@@

The previous work on js-scala \cite{Kossakowski12_JsDESL} showed how to make asynchronous programming more
convenient by making asynchronous calls looking like synchronous calls (\ie{} returning a value instead of taking a
callback as parameter). This work was based on the Scala continuations compiler plugin. We claim that, although this
programming model removes the “callback hell”, it can make the code hard to reason about because there is no explicit
distinction between synchronous and asynchronous parts.

We propose a DSL that explicitly reflects the asynchronous nature of computations in their types and provides methods
turning them into first-class citizens. The DSL is monadic so we can solve the callback hell problem thanks to the
Scala \code{for} notation.

For instance, listing \ref{async-first-class} shows how the listing \ref{async-js} can be re-written using our DSL.
The \code{createVertex} function now returns an asynchronous value instead of taking a callback as
parameter. Then, the \code{for} expression allows us to get the vertex, when available, and to insert it on the user
interface. By making the \code{createVertex} function return an asynchronous value instead of taking a callback as
a parameter, the code is easier to modularize into loosely coupled parts.

@@

The listing \ref{async-no-callback} translates the callback hell example (listing \ref{async-callback-hell}) using
our DSL. The \code{for} notation can intuitively be thought of as a sequencing notation: whenever the response of the
first Ajax request is available, the next statement will be executed, and so on. There is no nested callbacks and the
order of execution is directly reflected by the order of statements.

@@

In this section we show how we can define a template engine as an embedded DSL with minimal effort. This template
engine is statically typed and able to insert dynamic content in a safe way. It provides a powerful expression
language, requires no extra compilation step and can be used on both client-side and server-side.

@@

So the only remaining work consists in defining the DSL vocabulary to define DOM nodes. We provide a \code{tag}
function to define a tag and a \code{text} function to define a text node.

\begin{figure}
\begin{lstlisting}[label=forest-hello,caption=DOM definition DSL]
def vertexDom(v: Rep[Vertex]) =
    tag("g", "class"->"vertex",
             "transform"->("translate("+v.x+","+v.y+")"))(
        tag("rect", "width"->v.width, "height"->v.height)(),
        tag("text", "width"->v.width, "height"->v.height)(
            text(v.content)
        )
    )
\end{lstlisting}
\end{figure}

Listing \ref{forest-hello} uses our DSL and generates a code equivalent to listing \ref{dom-api}. The readability has
been highly improved: nesting tags is just like nesting code blocks, HTML entities are
automatically escaped in text nodes, developers have the full computational power of Scala to inject dynamic data and
DOM fragments definitions are written using functions so they compose just as functions compose. These benefits come
with no performance loss because the DSL generates code building DOM fragments by using the native JavaScript API.

@@

Our DSL is equivalent to a template engine with Scala as the expression language. Making it usable on both server and
client sides was surprisingly as simple as defining another code generator for the DSL, producing Scala code.

@@

Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
\emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
the objects. However, modifying existing object prototypes is considered bad
practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
registering supported data types at the beginning of the program execution, as described in section 2.4.3
of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.

@@

In summary, the compiled embedded approach may require more work to define custom data types but gives control on
the way these data types will be encoded for each target platform (\eg{} the JavaScript code generator uses native
arrays to encode the lists of the list manipulation DSL).

@@

Furthermore, a code generator can be tweaked by defining a sub-trait specializing some of its methods. This feature
can be used to handle cross-browser incompatibilities. For instance we have written code generators specializing the
output for some statement nodes in order to handle Internet Explorer incompatibilities.

@@

This section presents relevant implementation details of the js-scala DSL\footnote{The complete implementation is
available at \href{http://github.com/js-scala}{http://github.com/js-scala}}.

@@

The \emph{ad hoc} polymorphism feature of our DSL is directly borrowed from Scala’s typeclasses encoding but the type
coercion feature required to define a type-level DSL to express constraints. Numeric operations take an implicit
value of type \code{(A \~ B) ~> C}. This type is actually a syntactic sugar for the type \code{\~>[\~[A, B], C]}
(Scala supports an infix notation for type constructors of arity two). The \code{\~\>} type also takes two value
parameters, saying how to obtain a value of the target type for each operand. Here is the definition of the
\code{\~\>} type:

@@

  def option_isEmpty[A](o: Rep[Option[A]]): Rep[Boolean]
  def option_foreach[A](o: Rep[Option[A]],
      f: Rep[A] => Rep[Unit]): Rep[Unit]
}
// Intermediate representation
trait OptionOpsExp extends OptionOps { this: EffectExp =>
  def option_isEmpty[A](o: Exp[Option[A]]) = OptionIsEmpty(o)
  def option_foreach[A](o: Exp[Option[A]],
      f: Exp[A] => Exp[Unit]) = {
    val a = fresh[A]
    val b = reifyEffects(f(a))
    reflectEffect(OptionForeach(o, a, b), summarizeEffects(b).star)
  }

@@

We limited the DSL to two operations (\code{foreach} and \code{isEmpty}) for the sake of brevity. The
\code{OptionOps} trait defines the user facing DSL, it enriches \code{Rep[Option[_]]} values by providing an implicit
conversion to \code{OptionOpsCls} that defines the \code{isEmpty} and \code{foreach} methods that respectively tests
the emptiness of a value and derefences a value if not empty. These methods provide a convenient syntax to use
\code{Rep[Option[_]]} values and delegate to the \code{option_isEmpty} and \code{option_foreach} operations,
respectively, that are left abstract. The \code{OptionOpsExp} trait implements the \code{option_isEmpty} and
\code{option_foreach} methods. The \code{option_isEmpty} implementation simply returns a statement node of type
\code{OptionIsEmpty}, but the \code{option_foreach} implementation is more subtle because it has to track the
dependencies of its function parameter \code{f} (for instance, side-effects performed within the \code{f} function
must be reflected by the returned \code{OptionForeach} node).

The principle is the same for all DSLs: first, provide an abstract trait with the DSL vocabulary, then implement each
DSL construct to return a reified intermediate representation of itself. Sometimes you have the opportunity
to perform some optimizations at the evaluation level such as constants folding. For instance, the DOM definition DSL
builds trees consisting of a root node and eventually several children. If these children are known at evaluation
time we can build a node having a \code{List[Rep[Node]]} value as children instead of a \code{Rep[List[Node]]} so the
generated code will be able to inline the loop adding the children to the parent node. Listing \ref{forest-impl}
shows the relevant part of code checking if the \code{children} parameter of the \code{forest_tag} method is a
constant list (the \code{ListNew} statement node represents constant lists) and building different intermediate
representations according to this test.

\begin{figure}
\begin{lstlisting}[caption=Optimized intermediate representation for the DOM definition DSL,label=forest-impl]
trait ForestExp extends Forest with EffectExp { this: ListOpsExp =>
  def forest_tag(name: String, attrs: Map[String, Exp[String]],
      children: Exp[List[Node]]) = children match {
    case Def(ListNew(children)) =>
      Tag(name, Left(children.toList), attrs)
    case _ =>
      Tag(name, Right(children), attrs)
  }
  case class Tag(name: String,
      children: Either[List[Exp[Node]], Exp[List[Node]]],
      attrs: Map[String, Exp[String]]) extends Def[Node]
}
\end{lstlisting}
\end{figure}

@@

Code generation consists in traversing the statement nodes produced by the program evaluation according to their
dependencies and to emit the code corresponding to each statement. LMS already sorts the statements graph so DSL
authors just need to say how to emit code for each statement node of their DSL. Listing \ref{option-codegen} shows
such a code generator for the \code{null} reference handling DSL. The \code{emitNode} method handles
\code{OptionIsEmpty} and \code{OptionForeach} nodes. In the case of the \code{OptionIsEmpty} node, it simply
generates an expression testing if the value is \code{null}, in the case of the \code{OptionForeach} node, it wraps
the code block dereferencing the value within a \code{if} checking that the value is not \code{null}.

\begin{figure}
\begin{lstlisting}[caption=Null reference handling DSL code generator,label=option-codegen]
trait JSGenOptionOps extends JSGenEffect {
  val IR: EfffectExp with OptionOpsExp
  import IR._
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
    rhs match {
      case OptionIsEmpty(o) =>
        emitValDef(sym, quote(o) + " === null")
      case OptionForeach(o, a, b) =>
        stream.println("if (" + quote(o) + " !== null) {")
        emitValDef(a, quote(o))
        emitBlock(b)
        stream.println("}")
      case _ =>
        super.emitNode(sym, rhs)
    }
}
\end{lstlisting}
\end{figure}
